---
title: "R-code for 'A time-heterogeneous D-vine copula model for unbalanced and unequally spaced longitudinal data' "
author: "*Md Erfanul Hoque, Elif F. Acar and Mahmoud Torabi*"
output: pdf_document
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This script generates a random samples of repeated measurement data from the time-heterogeneous D-vine model (HET--P) presented in Section 4 of *A time-heterogeneous D-vine copula model for unbalanced and unequally spaced longitudinal data* by Hoque, Acar and Torabi (2020) and illustrates the associated R functions that implement HET--P model to estimate the model parameters and conditional quantiles.

# Load required packages
```{r, message=FALSE, warning=FALSE}
library(here) # to specify the folder where something is located in
library(VineCopula) # required to work with D-vine copulas
```

```{r, echo=FALSE}
# Load the auxiliary functions located in 'src' folder which is
# needed to generate data
source(here::here("src", "LDvine.aux.R"))
```

# Simulation Example: Time-heterogeneous D-vine copula under DGP4

```{r,cache=TRUE}
# Data example used for DGP 4 (HET--P) in manuscript (Section 4).
n <- 250 # sample size
d <-  5 # dimension

# Decay Parameter values
beta0 <- c(0, 0.9,0.53,0.34,0.06, 0,0,0.68,0.21,0.15,
           0,0,0,0.27,0.09, 0,0,0,0,0.11, 0,0,0,0,0)
beta0 <- matrix(beta0, d, d)

beta <- c(0,1.4,0.82,0.63,0.21, 0,0,0.94,0.78,0.42,
          0,0,0,0.69,0.28, 0,0,0,0,0.36, 0,0,0,0,0)
beta <- matrix(beta, d, d)

# Generate time variable
set.seed(20192020)
Time <- t(sapply(1:n, function(i) Longt.Sim(d)))

# Calculate Kendall's Tau
Tau.Array <- sapply(1:n, function(i) LDVine.Tau(beta0,beta, Time[i,]), 
                                                  simplify="array")
# Copula families in D-vine matrix 
Family <- c(0, 1, 1, 4, 4, 
            0, 0, 5, 3, 3,
            0, 0, 0, 5, 5,
            0, 0, 0, 0, 1, 
            0, 0, 0, 0, 0)
Family <- matrix(Family, nrow=d, ncol=d)
  
# Calculate copula parameters
Par.Array <- sapply(1:n, function(i) LDVine.Par(Family, Tau=Tau.Array[,,i]),
                                                  simplify="array") 
# Generate 5 dimensional balanced data
set.seed(20192020)
Full.data <- data.frame(t(sapply(1:n, function(i)  LDVine.Sim(d, Family, 
                                                        Par.Array[,,i]))))
colnames(Full.data) <- c("U1","U2","U3","U4","U5")

# Save the generated data in "results" folder 
save(Full.data,file=here::here("data","DGP4-n250.Rdata"))

###########################################################################
# Obtain unbalanced data from the balanced data
###########################################################################
prob <- c(20, 20, 15, 40)/100 # prob of d_i, d_i=2,3,4,5

# Sample lengths of observations
set.seed(12345) 
n.event <- sample(x = 2:d, size = n, replace = TRUE, prob = prob)
  
# Unbalanced/reduced data
U.RM <- matrix(NA, n, d)
  
# Reduced data length 
for (i in 1:n){
  U.RM[i, 1:n.event[i]] <- unlist(Full.data[i, 1:n.event[i]])
}
  
## Proportion of individual with at least j meaurements
ind.num<- c()
for (i in 1:d){
  ind.num[i]=sum(!is.na(U.RM[,i]))/n
}
ind.num

# Save the unbalanced data in "results" folder 
save(U.RM,file=here("data","Unbal-DGP4-n250.Rdata"))
```

# D-vine copula model fitting
We fit a time-homogeneous D-vine copula model (HOM--P) and a time-heterogeneous D-vine copula model (HET--P) to the data.

### Load copula data to fit D-vine copula models
```{r}
U.data<- U.RM # unobserved measurements are denoted by 'NA'
dim(U.data)
head(U.data) # show the copula data structure
```

### Fit time-homogeneous D-vine copula model (HOM--P) to copula data

```{r}
# Calculate Kendall's tau values
Tau.Array.True <- sapply(1:n, function(i) LDVine.Tau(beta0, beta, 
                                              Time[i,]), simplify="array")
Tau=apply(Tau.Array.True,1:2,mean) #average tau

# Calculate copula parameter values
Par <- matrix(0, d, d)
Par2 = matrix(0,nrow=d,ncol=d)
 for (i in 2:d) {
   for (j in 1:(i-1)) {
     Par[i, j] <- BiCopTau2Par(Family[i, j], Tau[i, j])
   }
 }
# D-vine matrix with fixed order
 elements <- 1:d
  Mat <- diag(1:d)
  for (i in 2:d){
    Mat[d+2-i, 1:(d+1-i)] <- elements[i:d]
  }

# Construct the RVM matrix to fit HOM-P 
RVM <- VineCopula::RVineMatrix(Mat, Family, Par, Par2)  

# Fit HOM-P with fixed order 1:5
Dvine.HOM = RVineSeqEst(data = U.data, RVM=RVM)
# The above warning only confirms that the data are unbalanced.
Dvine.HOM

# Copula log-likelihood value for HOM-P model
Cloglik_HOMByPair<- sum(Dvine.HOM$pair.logLik);Cloglik_HOMByPair

# Store the copula parameters and families for quantile prediction
par<-Dvine.HOM$par  
parset.HOM <-list(c(par[5,1],par[5,2],par[5,3],par[5,4]),
               c(par[4,1],par[4,2],par[4,3]),
               c(par[3,1],par[3,2]),
               c(par[2,1]))
fam<-Dvine.HOM$family
familyset.HOM <- list(c(fam[5,1],fam[5,2],fam[5,3],fam[5,4]),
                   c(fam[4,1],fam[4,2],fam[4,3]),
                   c(fam[3,1],fam[3,2]),
                   c(fam[2,1]))
```

### Fit time-heterogeneous D-vine copula model (HET--P) to copula data

To fit the HET--P model, we first calculate time interval between two measurements. The fitted model  gives the estimated values of the decay parameters for each pair copula.

```{r, echo=FALSE}
## FUNCTIONS EVALUATION
# Load the auxiliary functions located in 'src' folder which are 
# needed to fit the HET-P model
source(here::here("src", "LBiCopFit.R"))
source(here::here("src", "LBiCopLLik.R"))
```

```{r, warning=FALSE}
family.set <- familyset.HET<- familyset.HOM
# Creating the Time variable for the analysis of HET-P 
Visits <- data.frame(Time)
colnames(Visits)<-c("visit 1","visit 2","visit 3","visit 4","visit 5")
head(Visits)
## BEGIN fitted model 
# Calculate the time intervals for Tree 1
t <- sapply(1:(d-1), function(j) Visits[,j+1] - Visits[,j])  

# Estimation of model parameters of pair copulas in Tree 1
fit<- sapply(1:(d-1), function(j) LBiCopFit(u1=U.data[,j], u2=U.data[,j+1], 
                        x=t[,j],family=family.set[[1]][j],repar="pearson",
                        start = c(0.01,0.01)))

fit.beta.T1<- sapply(1:(d-1), function(j) c(fit[1,j],fit[2,j])) 
ll.T1<-sum(unlist(fit[3,])) #loglikelihood value in Tree 1

# Estimate copula parameters using fitted copulas
tau.T1<- sapply(1:(d-1), function(j) LDVine.rho2tau(unlist(fit.beta.T1[,j]),
                                                    t[,j]))
par.T1<- sapply(1:(d-1), function(j) BiCopTau2Par(family.set[[1]][j], 
                                                  tau.T1[,j]))
# Obtain the pseudo-conditional marginals for Tree 2
vv<-list()
for (j in 1:(d-2)){
  vv[[j]] <- BiCopHfunc2(u1=U.data[,j], u2=U.data[,j+1],par.T1[,j],
                                      family = family.set[[1]][j])
  vv[[(d-2)+j]] <- BiCopHfunc2(u1=U.data[,j+2], u2=U.data[,j+1],par.T1[,j+1],
                                      family =family.set[[1]][j+1])
}
# Calculate the time intervals for Tree 2
t<- sapply(1:(d-2), function(j) Visits[,j+2] - Visits[,j])  

# List of available conditional data needed for estimation in Tree 2
condata1 <- list(vv[[1]],vv[[4]],vv[[2]],vv[[5]],vv[[3]],vv[[6]])

# Estimation of model parameters of pair copulas in Tree 2
fit<- sapply(1:(d-2), function(j) LBiCopFit(u1 = condata1[[2*j-1]], 
                u2 = condata1[[2*j]], x=t[,j],family=family.set[[2]][j],
                repar="pearson",start = c(0.01,0.01)))

fit.beta.T2<- sapply(1:(d-2), function(j) c(fit[1,j],fit[2,j])) 
ll.T2<-sum(unlist(fit[3,])) #loglikelihood value in Tree 2

# Estimate copula parameters using fitted copulas
tau.T2<- sapply(1:(d-2), function(j) LDVine.rho2tau(unlist(fit.beta.T2[,j]),
                                                    t[,j]))
par.T2<- sapply(1:(d-2), function(j) BiCopTau2Par(family.set[[2]][j], 
                                                  tau.T2[,j]))
# Obtain the pseudo-conditional marginals for Tree 3
vvv<-list()
for (j in 1:(d-3)){
  vvv[[j]] <- BiCopHfunc2(u1=condata1[[2*j-1]], u2=condata1[[2*j]],
                                 par.T2[,j],family = family.set[[2]][j])
  vvv[[(d-3)+j]] <- BiCopHfunc2(u1=condata1[[2*j+2]], u2=condata1[[2*j+1]],
                                 par.T2[,j+1],family = family.set[[2]][j+1])
}
# Calculate the time intervals for Tree 3
t<- sapply(1:(d-3), function(j) Visits[,j+3] - Visits[,j])  

# List of available conditional data needed for estimation in Tree 3
condata2 <- list(vvv[[1]],vvv[[3]],vvv[[2]],vvv[[4]])

# Estimation of model parameters of pair copulas in Tree 3
fit<- sapply(1:(d-3), function(j) LBiCopFit(u1 = condata2[[2*j-1]], 
                u2 = condata2[[2*j]], x=t[,j], family=family.set[[3]][j],
                repar="pearson",start = c(0.01,0.01)))

fit.beta.T3<- sapply(1:(d-3), function(j) c(fit[1,j],fit[2,j])) 
ll.T3<-sum(unlist(fit[3,])) #loglikelihood value in Tree 3

# Estimate copula parameters using fitted copulas
tau.T3<- sapply(1:(d-3), function(j) LDVine.rho2tau(unlist(fit.beta.T3[,j]),
                                                    t[,j]))
par.T3<- sapply(1:(d-3), function(j) BiCopTau2Par(family.set[[3]][j], 
                                                  tau.T3[,j]))
# Obtain the pseudo-conditional marginals for Tree 4
vvvv<-list()
for (j in 1:(d-4)){
  vvvv[[j]] <- BiCopHfunc2(u1=condata2[[2*j-1]], u2=condata2[[2*j]],
                                 par.T3[,j],family = family.set[[3]][j])
  vvvv[[(d-4)+j]] <- BiCopHfunc2(u1=condata2[[2*j+2]], u2=condata2[[2*j+1]],
                                 par.T3[,j+1],family = family.set[[3]][j+1])
}

# Calculate the time intervals for Tree 4
t<- sapply(1:(d-4), function(j) Visits[,j+4] - Visits[,j])  

# List of available conditional data needed for estimation in Tree 4
condata3 <- list(vvvv[[1]],vvvv[[2]])

# Estimation of model parameters of pair copulas in Tree 4
fit<- sapply(1:(d-4), function(j) LBiCopFit(u1 = condata3[[2*j-1]],
               u2 = condata3[[2*j]], x=t[,j], family=family.set[[4]][j],
               repar="pearson",start = c(0.01,0.01)))

fit.beta.T4<- sapply(1:(d-4), function(j) c(fit[1,j],fit[2,j])) 
ll.T4<-sum(unlist(fit[3,])) #loglikelihood value in Tree 4

# Estimate copula parameters using fitted copulas
tau.T4<- sapply(1:(d-4), function(j) LDVine.rho2tau(unlist(fit.beta.T4[,j]),
                                                    t[,j]))
par.T4<- sapply(1:(d-4), function(j) BiCopTau2Par(family.set[[4]][j], 
                                                  tau.T4[,j]))
## END fitted model 

# Store the parameter estimates
beta.fit <- list(fit.beta.T1,fit.beta.T2,fit.beta.T3,fit.beta.T4)

# Store the copula parameters and families for quantile prediction
par.fit.HET <- list(par.T1,par.T2,par.T3,par.T4)
familyset.HET <- family.set

# Copula log-likelihood value for HET-P model
Cloglik_HETByPair <- ll.T1+ll.T2+ll.T3+ll.T4; Cloglik_HETByPair
```

# Implementation of Conditional Quantile Prediction

We select a subject with 5 measurements. We pretend that the selected subject has only 4 measurements rather than 5 measurements and predict the conditional $5\%$-, $50\%$- and $95\%$-quantiles in copula scale for its fifth measurement based on the first four measurements. 

```{r, echo=FALSE}
# Load the auxiliary functions located in 'src' folder which is
# needed to do the D-vine copula quantile prediction
source(here::here("src", "LDvineQuantile.R"))
```

```{r}
# id of selected individual and corresponding vector of measurements
id.sel <- 15
U.sel <- U.data[id.sel,]
U.sel
time.sel <- Visits[id.sel,]

# Quantile levels to be estimated
alpha.level <- c(0.05, 0.5, 0.95)
```

For the D-vine copula models HOM--P and HET--P, we use the copula data to calculate the D-vine quantile of the fifth measurement. 

```{r}
udata = rbind(U.sel[1:4])
```

### Time-homogeneous D-vine copula model (HOM--P)

```{r}
# Calculate the quantiles on the coupla scale
u_quant.p5 <-  DvineQuant (p = alpha.level[1], u = udata, 
                    family.set = familyset.HOM, par.set = parset.HOM)
u_quant.p50 <- DvineQuant (p = alpha.level[2], u = udata, 
                    family.set = familyset.HOM, par.set = parset.HOM)
u_quant.p95 <- DvineQuant (p = alpha.level[3], u = udata, 
                    family.set = familyset.HOM, par.set = parset.HOM)
U.quant.HOM<-c(u_quant.p5,u_quant.p50,u_quant.p95)
U.quant.HOM
# For a given marginal distribution, we can easily transform the estimated
# quantiles from the copula scale to the original scale
```

### Time-heterogeneous D-vine copula model (HET--P)

```{r}
# Estimated copula parameters from fitted HET-P model 
parset.HET <- sapply(1:(d-1), function(j) par.fit.HET[[j]][id.sel,])

# Calculate the quantiles on the copula scale
u_quant.p5 <- DvineQuant (p = alpha.level[1], u = udata, 
                       family.set = family.set, par.set = parset.HET)
u_quant.p50 <- DvineQuant (p = alpha.level[2], u = udata, 
                       family.set = family.set, par.set = parset.HET)
u_quant.p95 <- DvineQuant (p = alpha.level[3], u = udata, 
                       family.set = family.set, par.set = parset.HET)
U.quant.HET<-c(u_quant.p5,u_quant.p50,u_quant.p95)
U.quant.HET

# For a given marginal distribution, we can easily transform the estimated 
# quantiles from the copula scale to the original scale

# Save the copula prediction results in "results" folder 
U.quant<-rbind(U.quant.HOM,U.quant.HET)
rownames(U.quant)<-c("HOM","HET")
colnames(U.quant)<-c("5%","50%","95%")
U.quant
U.sel

write.table(round(U.quant,2),file=here("results","copula.quant.txt"))
write.table(U.sel,file=here("results","copula.data.txt"))
```
